A self-consistency check for unitary propagation of Hawking quanta

### 2016-12-18 Draft_BH_shells.tex

While formulating the information paradox, it is customary to take the large black hole limit.
Technically, it means that we first select a fixed length scale $l$ that is much longer than the UV-cutoff scale of the theory, and study a near-horizon region of this size.
Such a region will be almost identical to a region of the same size in global Minkowski space, since all locally measurable curvature effect is suppressed by $(l/M)$.
If we make the black hole arbitrarily large, $(l/M)$ goes to zero, the for all the field theory modes within this region, we are confident that they start in local vacuum states of QFT as if they live in Minkowski space.
This large black hole limit is the key argument to eliminate many na\"ive solutions to the information paradox.

The main result in this paper shows that there is a competing effect which stops us from taking this convenient limit.
In order for field theory modes from the above region to propagate out and become asymptotic quanta, they must go through geometric uncertainties, like the double-shells in our example, at the distance scale $l$ from the horizon.
If we again fix $l$ and increase $M$, our result in Eq.~(\ref{fin-result}) shows that the unitary propagation to asymptotic Hawking quanta becomes less and less trustworthy. 
Since the dimensionless ratio $(l/M)$ is the only relevant physical parameter that comes into the scaling both effects, there seems to be no middle-ground that can make both unitarity and local-Minkowski-ness arbitrarily reliable.
We think such competition is a good sign and will take us closer to the solution of information paradox.

Note that our argument will fail if the combination $(\sigma\Delta l)$ is exactly zero instead of a fixed finite number.
Namely, if there is absolutely no geometric uncertainty, then the geometry is indeed a completely classical system that does not entangle with the quantum field.
We would like to argue that in the context of QFT in curved spacetime, enforcing $\Delta g_{\mu\nu}=0$ is quite unreasonable.
After all, we are feeding the expectation value of the stress-tensor-operator, $\langle T_{\mu\nu}\rangle$, to the Einstein equation. 
As long as we are not in an eigenstate of the local stress density, $\langle\Delta T_{\mu\nu}\rangle$ will be nonzero, and there is no reason why this does not lead to an uncertainty in the geometry.
One can even use the uncertainty in the energy flux of Hawking radiation itself to argue that there has to be an uncertainty in geometry.
\footnote{Although, if one simply blue shifts the actual out-going Hawking radiation back to the hear-horizon zone, one gets a contribution of $(\sigma\Delta l)$ that depends on $M$.
That is because the QFT modes escaping the angular-momentum barrier are mostly in $S$-waves.
This $S$-wave energy flux gets diluted by area, resulting $(\sigma\Delta l)\propto M^{-2}$.
If this is the only contribution to $(\sigma\Delta l)$, such dependence exactly stops Eq.~(\ref{fin-result}) from diverging in the large black hole limit, which undermines our main argument in this paper.
Nevertheless, it violates equivalence principle for the local property $(\sigma\Delta l)$ to depend on the size of the black hole.
It will be interesting to explicitly compute $\langle\Delta T_{\mu\nu}\rangle$ for the actual states of near-horizon QFT modes to see whether they uphold equivalence principle.
Note that higher angular modes are all excited near the horizon.
They are trapped by the angular momentum barrier thus do not contribute to the asymptotic flux, but they should still contribute to local quantities.
That is why the $S$-wave contribution might be a dramatic underestimation, and the actual $(\sigma\Delta l)$ can still respect equivalence principle.
If that is true, our arguments will hold.
We will leave that particularly involved QFT calculation to future works.
 }

>>>>

While formulating the information paradox, it is customary to take the large black hole limit.
Technically, it means that we first select a fixed length scale $l$ that is much longer than the UV-cutoff scale of the theory, and study a near-horizon region of this size.
Such a region will be almost identical to a region of the same size in global Minkowski space, since all locally measurable curvature effect is suppressed by $(l/M)$.
If we make the black hole arbitrarily large, $(l/M)$ goes to zero, then for all the field theory modes within this region, we are confident that they start in local vacuum states of QFT as if they live in Minkowski space.
This large black hole limit is the key argument to eliminate many na\"ive solutions to the information paradox.

The main result in this paper shows that there is a competing effect which stops us from taking this convenient limit.
In order for field theory modes from the above region to propagate out and become asymptotic quanta, they must go through geometric uncertainties.
These geometric uncertainties are like the double-shells in our example, and we only care about those that an out-going quantum has to pass through---those at distance scales larger than $l$ from the horizon.
If we again fix $l$ and increase $M$, our result in Eq.~(\ref{fin-result}) shows that the unitary propagation to asymptotic Hawking quanta becomes less and less trustworthy.

The dimensionless ratio $(l/M)$ is the only relevant physical parameter that comes into the scaling of both of the above effects.
The suppressing of curvature effects prefers this ratio to be large, but the validity of unitary propagation prefers this to be small.
Therefore, we cannot make both unitarity and local-Minkowski-ness arbitrarily reliable.
We think such competition is a good sign and will take us closer to the solution of information paradox.

Our argument will fail if the geometric uncertainty is exactly zero instead of a fixed finite number.
With an exactly zero uncertainty, the geometry is indeed a completely classical system that does not entangle with the quantum field
\footnote{Alternatively, if $(\sigma\Delta l)$ is really small, there might be an intermediate value of $(l/M)$ that makes both the curvature effects small and a unitarity propagation sufficiently trustworthy.
A more involved discussion is required for such possibility and we postpone that to future work.}.
We would like to argue that in the context of QFT in curved spacetime, enforcing $\Delta g_{\mu\nu}=0$ is quite unreasonable.
After all, we are feeding the expectation value of the stress-tensor-operator, $\langle T_{\mu\nu}\rangle$, to the Einstein equation. 
As long as we are not in an eigenstate of the local stress density, $\langle\Delta T_{\mu\nu}\rangle$ will be nonzero, and there is no reason why this does not lead to an uncertainty in the geometry.
One can even use the uncertainty in the energy flux of Hawking radiation itself to argue that there has to be an uncertainty in geometry.

If one simply blue-shifts the actual out-going Hawking radiation energy flux back to the hear-horizon zone, one gets a contribution of $(\sigma\Delta l)\propto M^{-2}$.
This would have nullified our argument since Eq.~(\ref{fin-result}) will no longer depend on $M$
\footnote{One can still take the face value of Eq.~(\ref{fin-result}) and claim that a propagation from the Planckian, near-horizon region loses unitarity.
This means solving the information problem with something similar to a firewall \cite{AMPS}:  a near-horizon observer can see violation of equivalence principle in scales arbitrarily smaller than the Schwarzschild radius.}.
However, such an $M$-dependent value of geometric uncertainty explicitly violates the equivalence principle.
Using such value in our analysis would be a technical mistake.
The value of energy flux that can escape the angular momentum barrier are mostly in $S$-waves and it depends explicitly on $M$.
However, the geometric uncertainty near the horizon are also affected by all those modes trapped inside the angular momentum barrier.
Inferring a near-horizon geometric uncertainty solely from the asymptotic flux is a dramatic underestimation.
In fact, instead of using the Rindler basis which is natural for out-going modes to calculate the energy flux, one can use the locally Minkowski mode basis to calculate the local stress tensor.
It should then be obvious that $\langle\Delta T_{\mu\nu}\rangle$ upholds equivalence principle, which demands an $M$-independent geometric uncertainty
\footnote{Although, we are not advocating that such calculation directly gives us the value of $(\sigma\Delta l)$.
That is because using the expectation value of $T_{\mu\nu}$ in the Einstein equation is itself an assumption.
The actual value one can use is related to the decoherence problem which we would like to circumvent for now.
Our argument is valid as long as the geometric uncertainty is nonzero and obeys equivalence principle.}.
